Predicting Bike Usage at UofT St. George Campus with Bayesian Poisson Regression

### 2025-01-17 paper/paper.qmd
Typos fixed
Typos fixed

---
title: "Predicting Bike Usage at UofT St. George with Bayesian Poisson Regression"
subtitle: "Daily usage grows by one every 20 days, drops to between 0 and 3 in winter, with 79% occurring between 8 AM and 8 PM"
author: 
  - Haowei Fan
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Bikeshare-Forecast.git"
date: today
date-format: long
abstract: "University of Toronto's St. George campus students and staff frequently encounter challenges in locating available bike-sharing parking spots upon arrival and bikes upon leaving [@elassi2017bike], underscoring the necessity for accurate predictions of future bike-sharing demand to optimize campus commuting infrastructure. This study employs Bayesian Poisson regression to predict the utilization of bike-sharing stations at 27 locations across campus for specific years, months, and four-hour intervals of the day. The results show that daily bike-sharing usage on campus increases by one every 20 days but decreases to between 0 and 3 in winter months, and 79% of peak usage consistently observed between 8:00 AM and 8:00 PM daily. Specifically, on September 26, 2025, between 4:00 and 8:00 AM, the usage of 8 stations is projected to increase by 3 bikes, 14 stations by 2 bikes, 3 stations by 1 bike, while 1 station will remain unchanged, and 1 station will see a decrease by 1 bike, compared to the same timeframe in 2024."
format: pdf
toc-depth: 2    
number-sections: true
toc: true           
toc-title: "Table of Contents" 
bibliography: references.bib
---

@@

## Overview
Bike-sharing services have experienced growth in the Toronto region, with ridership rising dramatically from approximately 665,000 trips in 2015 to over 4.5 million in 2022, according to a study by the University of Toronto's School of Cities[@liu2023bike]. This increase highlights the need for analyses to optimize bike-sharing systems and guide policy decisions that can further enhance urban mobility[@xiamen]. While city-wide studies provide suggestions, research from Hangzhou, China, has demonstrated that bike-sharing usage patterns can vary significantly between urban areas and university campuses[@hangzhou]. These differences suggest that general urban studies conducted in the Toronto region may not fully capture the unique dynamics of campus-based bike-sharing systems. Therefore, to address this unexplored areas in the literature, this study focuses on the University of Toronto St. George Campus as a case study, analyzing bike-sharing usage within the campus in Toronto to provide a detailed understanding of its role in campus mobility and inform policies tailored to similar settings.

@@

## Overview
The dataset used in this study is named "Bike Share Toronto Ridership Data" [@bikesharedata]. It comes from opendatatoronto [@opendata], uploaded by Toronto Parking Authority[@greenp] and collected by Bike Share Toronto[@bike]. It records every bike-sharing usage in the Toronto area from 2015 to September 30, 2024, with a total of 28,017,329 records. The variables included in the data differ across years, but they all contain the following variables: Trip ID, Trip Duration, Trip Start Station ID, Trip Start Time, Trip Start Station Location, Trip End Station ID, Trip End Time, Trip End Station Location, Bike ID, and User Type.

@@

**tidyverse**[@tidyverse]: Used for data wrangling, cleaning, and analysis, integrating multiple data manipulation packages.

**dplyr**[@dplyr]: Used for data frame operations such as filtering, sorting, and aggregation.

**arrow**[@arrow]: Used for reading and writing data in efficient formats like Parquet to speed up processing.

**stringr**[@stringr]: Used for handling and manipulating strings, allowing for text data cleaning and transformation.

**readr**[@readr]: Used for fast reading of CSV and other text formats.

**lubridate**[@lubridate]: Used for handling and converting date-time data, simplifying time-related data analysis.

**brms**[@brms]: Used for building and estimating Bayesian regression models for flexible data analysis.

**rstanarm**[@rstanarm]: Used for Bayesian regression modeling, helping to understand uncertainty better.

**bayesplot**[@bayesplot]: Used for visualizing posterior distributions and diagnostic plots of Bayesian models.

**ggplot2**[@ggplot2]: Used for creating various types of plots, supporting exploratory data analysis and result presentation.

**RColorBrewer**[@RColorBrewer]: Used for generating color palettes to create visually distinct and appealing plots.

**knitr**[@knitr]: Used for generating dynamic reports, integrating analysis results into documents.

**kableExtra**[@kableExtra]: Used for enhancing the presentation of tables, making tables in reports more visually appealing.

**scales**[@scales]: Used for formatting large numbers.

**reshape2**[@reshape]: Used for reshaping redata, particularly converting between wide and long formats

**testthat**[@tt]: Used to test whether the simulated data and analysis data meet the expected requirements.

**styler**[@styler]: Used to organize code formatting and code comments.

@@

Additionally, @fig-daily_usage shows an abnormal surge in usage from June to September 2024, which can be attributed to two main reasons. First, historical data indicates that usage typically spikes in August and September each year, likely due to increased demand from students returning to campus for the start of the academic year. Second, during the same period in 2024, Bike Share Toronto added 52 new stations, most of which were located near the St. George campus, significantly enhancing the network's coverage and convenience[@bike]. These two factors combined led to the appearance of anomalies in August and September 2024.

@@

# Model {#sec-model}
In the process of model development, this study utilized R [@citeR] and the following R packages:brms[@brms] and rstanarm[@rstanarm] were used for model construction. bayesplot[@bayesplot], tidyverse[@tidyverse], ggplot2[@ggplot2], knitr[@knitr], and kableExtra[@kableExtra] were used for verifying model assumptions and model validation.

@@

## Validation
In this section, the study selects "Galbraith Rd / King's College Rd" as the most representative station to validate the model. This station was chosen for several reasons: it is centrally located within the University of Toronto's St. George campus, it ranks among the top five stations in terms of data volume, and it has maintained the shortest average maintenance time every six months since its establishment[@bike].

>>>>

---
title: "Predicting Bike Usage at UofT St. George Campus with Bayesian Poisson Regression"
subtitle: "Daily usage grows by one every 20 days, drops to between 0 and 3 in winter, with 79% occurring between 8 AM and 8 PM"
author: 
  - Haowei Fan
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Bikeshare-Forecast.git"
date: December 14, 2024
date-format: long
abstract: "University of Toronto's St. George campus students and staff frequently encounter challenges in locating available bike-sharing parking spots upon arrival and bikes upon leaving [@elassi2017bike], underscoring the necessity for accurate predictions of future bike-sharing demand to optimize campus commuting infrastructure. This study employs Bayesian Poisson regression to predict the utilization of bike-sharing stations at 27 locations across campus for specific years, months, and four-hour intervals of the day. The results show that daily bike-sharing usage on campus increases by one every 20 days but decreases to between 0 and 3 in winter months, and 79% of peak usage consistently observed between 8:00 AM and 8:00 PM daily. Specifically, on September 26, 2025, between 4:00 and 8:00 AM, the usage of 8 stations is projected to increase by 3 bikes, 14 stations by 2 bikes, 3 stations by 1 bike, while 1 station will remain unchanged, and 1 station will see a decrease by 1 bike, compared to the same timeframe in 2024."
format: pdf
toc-depth: 2    
number-sections: true
toc: true           
toc-title: "Table of Contents" 
bibliography: references.bib
---

@@

## Overview
Bike-sharing services have experienced growth in the Toronto region, with ridership rising dramatically from approximately 665,000 trips in 2015 to over 4.5 million in 2022, according to a study by the University of Toronto's School of Cities [@liu2023bike]. This increase highlights the need for analyses to optimize bike-sharing systems and guide policy decisions that can further enhance urban mobility [@xiamen]. While city-wide studies provide suggestions, research from Hangzhou, China, has demonstrated that bike-sharing usage patterns can vary significantly between urban areas and university campuses [@hangzhou]. These differences suggest that general urban studies conducted in the Toronto region may not fully capture the unique dynamics of campus-based bike-sharing systems. Therefore, to address this unexplored areas in the literature, this study focuses on the University of Toronto St. George Campus as a case study, analyzing bike-sharing usage within the campus in Toronto to provide a detailed understanding of its role in campus mobility and inform policies tailored to similar settings.

@@

## Overview
The dataset used in this study is named "Bike Share Toronto Ridership Data" [@bikesharedata]. It comes from opendatatoronto [@opendata], uploaded by Toronto Parking Authority [@greenp] and collected by Bike Share Toronto [@bike]. It records every bike-sharing usage in the Toronto area from 2015 to September 30, 2024, with a total of 28,017,329 records. The variables included in the data differ across years, but they all contain the following variables: Trip ID, Trip Duration, Trip Start Station ID, Trip Start Time, Trip Start Station Location, Trip End Station ID, Trip End Time, Trip End Station Location, Bike ID, and User Type.

@@

**tidyverse** [@tidyverse]: Used for data wrangling, cleaning, and analysis, integrating multiple data manipulation packages.

**dplyr** [@dplyr]: Used for data frame operations such as filtering, sorting, and aggregation.

**arrow** [@arrow]: Used for reading and writing data in efficient formats like Parquet to speed up processing.

**stringr** [@stringr]: Used for handling and manipulating strings, allowing for text data cleaning and transformation.

**readr** [@readr]: Used for fast reading of CSV and other text formats.

**lubridate** [@lubridate]: Used for handling and converting date-time data, simplifying time-related data analysis.

**brms** [@brms]: Used for building and estimating Bayesian regression models for flexible data analysis.

**rstanarm** [@rstanarm]: Used for Bayesian regression modeling, helping to understand uncertainty better.

**bayesplot** [@bayesplot]: Used for visualizing posterior distributions and diagnostic plots of Bayesian models.

**ggplot2** [@ggplot2]: Used for creating various types of plots, supporting exploratory data analysis and result presentation.

**RColorBrewer** [@RColorBrewer]: Used for generating color palettes to create visually distinct and appealing plots.

**knitr** [@knitr]: Used for generating dynamic reports, integrating analysis results into documents.

**kableExtra** [@kableExtra]: Used for enhancing the presentation of tables, making tables in reports more visually appealing.

**scales** [@scales]: Used for formatting large numbers.

**reshape2** [@reshape]: Used for reshaping redata, particularly converting between wide and long formats

**testthat** [@tt]: Used to test whether the simulated data and analysis data meet the expected requirements.

**styler** [@styler]: Used to organize code formatting and code comments.

@@

Additionally, @fig-daily_usage shows an abnormal surge in usage from June to September 2024, which can be attributed to two main reasons. First, historical data indicates that usage typically spikes in August and September each year, likely due to increased demand from students returning to campus for the start of the academic year. Second, during the same period in 2024, Bike Share Toronto added 52 new stations, most of which were located near the St. George campus, significantly enhancing the network's coverage and convenience [@bike]. These two factors combined led to the appearance of anomalies in August and September 2024.

@@

# Model {#sec-model}
In the process of model development, this study utilized R [@citeR] and the following R packages:brms [@brms] and rstanarm [@rstanarm] were used for model construction. bayesplot [@bayesplot], tidyverse [@tidyverse], ggplot2 [@ggplot2], knitr [@knitr], and kableExtra [@kableExtra] were used for verifying model assumptions and model validation.

@@

## Validation
In this section, the study selects "Galbraith Rd / King's College Rd" as the most representative station to validate the model. This station was chosen for several reasons: it is centrally located within the University of Toronto's St. George campus, it ranks among the top five stations in terms of data volume, and it has maintained the shortest average maintenance time every six months since its establishment [@bike].
